---
title: "Git, GitHub & Ocean Acidification"
author: "<your name and u number>"
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This tutorial will draw on the skills you have already learnt on data visualisation, data wrangling and NHST to analyse a data set on ocean acidification effects on fish behaviour. However, we'll do that in the context of some new workflows that adopt Git and GitHub. 

Use this Rmarkdown file as your template to complete the tasks outlined within the instructions. There are questions asked in the instructions and here in the RMarkdown file. Answer these as best as you can to document your steps and thinking. There are also a series of coding exercises. 

Throughout we are also going to learn a bit more about how to format Rmarkdown documents. So, pay close attention to the formatting. We're also going to make use of the `bookdown` package now to render your Rmarkdown because, well, it's quite elegant in many ways. Before you start the tasks install the following packages:

```{r loadpacks, message=FALSE, results='hide'}
# Install a load of packages that we'll use. I'll show you a shortcut that I love to use. Try using the p_load function in the "pacman" package. p_load will execute both the install.packages and library commands in one shot so they only need to be used once to install pacman itself.
install.packages("pacman", repos = "http://cran.us.r-project.org")
library(pacman)

# Install bookdown for rendering because we'll need this. While we're at it, lets also install /load the tidyverse
p_load(bookdown, tidyverse, ggforce) # basically just list all the packages you want here

```

# *Task 1*

### Provide the link to your GitHub Repo 

While the tasks are meant to get you familiar with how GitHub works you should try and make regular commits as you work your way through. We want to see how often you're using Git. Below, within the `[]` brackets provide the link to your GitHub Repo. Just cut and paste it so we know what the repo name is and where it is. 

(My GitHub Repository)[]

By adding the link in the `[]` brackets it will hyperlink "My GitHub Repository". This is how to hyperlink in Rmarkdown documents.

# *Task 2*

This task involves cloning your repository on your computer. Where should we place this repository? While GitHub is somewhat of a cloud storage system because it's keeping record of your commits, files and changes in the cloud it is **not** a backup of your files (note that placing `**` around a word will bold it whereas '*' around a word will italicise it). This is important to recognise! 

>**Question 1**: Why should you not rely on *GitHub* as a backup?

```{r, answer1, echo=TRUE, eval = FALSE}
# Answer the question above here. You may want to explore further and then come back to your reasons as you familiarise yourself

# Use pound symbols to annotate your text.

```

I'd usually suggest placing your repo in a cloud storage system like, *OneDrive*, *Dropbox*, or *GoogleDrive*. These will ensure that all your files are backed up and *GitHub* will provide a record of the detailed history. 

Before moving on you might have noticed that the code block above has a name `answer1` and I've added an argument `echo = TRUE`. It's good to label code blocks with useful names to help you debug when rendering of your document goes wrong (when, not IF!). These names need to all be unique. There are a whole bunch of arguments that you can list within code chunks that tell each chunk what to do. You can look to the great [Rmarkdown cheetsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) for more information.

>**Question 2**: In the second code block called `answer1` if we change echo = TRUE to echo = FALSE what will happen to the code block `answer1` when you render it?

```{r answer2, echo=TRUE, eval = FALSE}
## Write your answer here....remember to use '#" in front of your text.
```

**Important**: Notice that `eval = FALSE` for many code chunks. When you complete the tasks. Make sure to change these to `TRUE` otherwise you will find your code does not run!

# *Task 3*

Hopefully you have now successfully cloned down your *GitHub* repo. This is your new working directory from which your project will develop. 

Hopefully you also created and/or downloaded the following files and added them to your cloned repo:

1) created a new RStudio project file in your working directory
2) download the `OA_activitydat_20190302_BIOL3207.csv` from Wattle and add this to a `data/` folder in your repo 
3) Add the template Rmarkdown file, `git_fishes_student.Rmd` that was provided on wattle. You'll use this file to work through the tasks and answer questions. 

This is a very good point to make sure that you understand relative path names fo coding. Wait, "relative what?" 

Your working directory can be considered the 'lowest' level of your project. The starting point, if you will. R needs to know where your files are that you want to load. Many people might be familiar with using the `setwd()` function to tell R where your files are that you need to load. This is problematic because everyone's path to a given working directory is different, and so, your code will not work on others' computers. Even across your own computers!

Lucky, RStudio project files take away the main issue by creating a project file that allows anyone to click on it to open the working directory to the same "spot". Having said that, if you start building structure (i.e., folders) in your working directory or repo than you need to understand how to navigate between folders to load and write files in the places that you want. 

Now that you have already added a `data/` folder to your working directory that you just cloned down think about how you would write some code to load in a data file that is within this folder. There are a few useful shortcuts you can use. For example, if you use `.` it means 'start from the current working directory'. That is the RStudio project file location if you click on a RStudio project. If you use `..` it means "start from the directory or folder one step back from the existing working directory".

>**Question 3**: **How would you write the path name to load in a data file from the data folder?**

Write your answer in the code chunk below:

```{r, loaddata, eval = FALSE, echo=TRUE}

path <- "replace this text with your relative path to a file in the 'data/' folder"
  
data <- read_csv(path)

```

Let's say that you added a new data folder that was inside an output folder (i.e., `output/data`) to your working directory. The purpose of this folder is to store and track your cleaned up data frame after you have done all your data wrangling and corrections. 

>**Question 4**: Using the `write_csv` function how would you write the path name to save this file to `output/data`?

```{r, write, eval = FALSE, echo=TRUE}

path <- "replace this text with your relative path"

write_csv(data, file = path)

```



