---
title: "Git, GitHub & Ocean Acidification"
author: "<your name and u number>"
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This tutorial will draw on the skills you have already learnt on data visualisation, data wrangling and NHST to analyse a data set on ocean acidification effects on fish behaviour. However, we'll do that in the context of some new workflows that adopt Git and GitHub. 

Use this Rmarkdown file as your template to complete the tasks outlined within the instructions. There are questions asked in the instructions and here in the RMarkdown file. Answer these as best as you can to document your steps and thinking. There are also a series of coding exercises. 

Throughout we are also going to learn a bit more about how to format Rmarkdown documents. So, pay close attention to the formatting. We're also going to make use of the `bookdown` package now to render your Rmarkdown because, well, it's quite elegant in many ways. Before you start the tasks install the following packages:

```{r loadpacks, message=FALSE, results='hide'}
# Install a load of packages that we'll use. I'll show you a shortcut that I love to use. Try using the p_load function in the "pacman" package. p_load will execute both the install.packages and library commands in one shot so they only need to be used once to install pacman itself.
install.packages("pacman", repos = "http://cran.us.r-project.org")
library(pacman)

# Install bookdown for rendering because we'll need this. While we're at it, lets also install /load the tidyverse
p_load(bookdown, tidyverse, ggforce) # basically just list all the packages you want here

```

# *Task 1*

### Provide the link to your GitHub Repo 

While the tasks are meant to get you familiar with how GitHub works you should try and make regular commits as you complete tasks. We want to see how often you're using Git. Below, within the `[]` brackets provide the link to your GitHub Repo. Just cut and paste it there so we know what the repo name is and where it is. 

(My GitHub Repository)[]

By adding the link in the `[]` brackets it will hyperlink "My GitHub Repository". This is how to hyperlink in Rmarkdown documents.

# *Task 2*

While GitHub is somewhat of a cloud storage system because it's keeping record of your commits, files and changes in the cloud it is **not** (note that placing `**` around a word will bold it whereas '*' around a word will italicise it) a backup of your files. This is important to recognise! 

>**Question 1**: Why should you not rely on *GitHub* as a backup?

```{r, answer1, echo=TRUE}
# Answer the question above here. You may want to explore further and then come back to your reasons as you familiarise yourself

# Use pound symbols to annotate your text.

```

I'd usually suggest placing your repo in a cloud storage system like, *OneDrive*, *Dropbox*, or *GoogleDrive*. These will ensure that all your files are backed up and *GitHub* will provide a record of the detailed history. 

> Stop here and try rendering your Rmarkdown to an html document. Have a look at how the document looks in the html rendered version. 

>**Question 2**: In the second code block called `answer1` try changing echo = TRUE to echo = FALSE. Re-render your document. What has changed? When you have answered this question in the code block below, change it back to echo = TRUE. 

```{r answer2, echo=TRUE}
## Write your answer here....remember to use '#" in front of your text.
```




