---
title: "Meta-analysis: Understanding and detecting publication biases"
date: "`r Sys.Date()`"
bibliography: ./bib/refs.bib
output: 
  bookdown::html_document2:
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, tidy = TRUE)
options(digits=2)
```

```{r klippy, echo=FALSE, include=TRUE, message=FALSE, warning=FALSE}
#install.packages("devtools")
remotes::install_github("rlesur/klippy")
klippy::klippy(tooltip_message = 'Click to Copy Code', tooltip_success = 'Done', position = 'right', color = "red")

# Load packages
pacman::p_load(metafor, flextable, tidyverse, orchaRd, pander, mathjaxr, equatags, vembedr)

# To use mathjaxr you need to run equatags::mathjax_install()
```

## **Introduction: Publication biases and what to do about them?**

Meta-analyst's have worked hard to develop tools that can be used to try and understand different forms of publication practices and biases within the scientific literature. Such biases can occur if studies reporting non-significant or opposite results to what is predicted are not found in systematic searches ['i.e., the 'file-drawer' problem; @Jennions2013]. Alternatively, biases could result from selective reporting or 'p-hacking'. 

Visual and quantitative tools have been developed try and identify and 'correct' for such biases on meta-analytic results [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Having said that, aside from working hard to try and incorporate 'gray literature' (unpublished theses, government reports, etc.) and working hard to include work done in non-English speaking languages, there is little one can truly due to counteract publication biases beyond a few simple tools. We cannot know for certain what isn't published in many cases or how a sample of existing work on a topic might be biased. Nonetheless, exploring the possibility of publication bias and its possible effects on conclusions is a core component of meta-analysis [@ODea2021]. 


## **Load the necessary R Packages**

```{r loadpacks, message=FALSE, results='hide'}
# Install a load of packages that we'll use. I'll show you a shortcut that I love to use. Try using the p_load function in the "pacman" package. p_load will execute both the install.packages and library commands in one shot so they only need to be used once to install pacman itself.
install.packages("pacman", repos = "http://cran.us.r-project.org")
library(pacman)

# Install bookdown for rendering because we'll need this. While we're at it, lets also install /load the tidyverse
p_load(bookdown, tidyverse, ggforce, flextable, latex2exp, png, magick, metafor) # basically just list all the packages you want here

```


## **References**

<div id="refs"></div>

<br>
