---
title: "Version Control, Git and GitHub"
author: "Daniel Noble"
date: "`r Sys.Date()`"
output: 
    bookdown::html_document2: 
      css: style.css
      toc: yes
      toc_depth: 4
      theme: cosmo
      number_sections: no
      toc_float: yes
      highlight: pygments
      fig_width: 4 
      fig_height: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vembedr)
```

```{r klippy, echo=FALSE, include=TRUE}
library(klippy)
klippy::klippy(
  lang = c("r", "markdown"),
  all_precode = FALSE,
  position = c("top", "right"),
  color = "darkred",
  tooltip_message = "Copy code",
  tooltip_success = "Copied!"
)
```

### **Learning objectives** {.tabset .tabset-fade .tabset-pills}

*   What is version control?
*   What is GitHub and why use it?
*   What's the difference between Public/Private repos?
*   How to set up a repo?
*   How to commit?
*   How to add collaborators?
*   How to use issues?
*   How to resolve conflicts?

<br>

#### Ground rules {.tabset .tabset-fade .tabset-pills}

Here are a few ground rules to keep things running smoothly:

1. Do your best to attempt each task on your own before asking for help from the instructors or your neighbour

2. When you get to a green box, stick the green sticky note to your laptop (or give yourself a green 'yes' tick on zoom)

3. If you need help, stick the red sticky note to your laptop (or raise your hand on zoom)
<br>

#### Getting un-stuck {.tabset .tabset-fade .tabset-pills}

Remember, no programmers know it all, so don't struggle alone too long, but before you ask for help do try:

* If you're looking for ideas of how to start:
  * googling for what you want to do (e.g. 'creating a repo github')
  * searching [GitHub Desktop Documentation](https://docs.github.com/en/desktop/installing-and-configuring-github-desktop/overview/getting-started-with-github-desktop) for what you want
  * use the amazing [Git Tutorial Videos](https://www.gitkraken.com/learn/git/tutorials)

If none of this works, ask your neighbour, use your red sticky note (or raised hand on zoom) to get help from a demonstrator. 
<br>

#### Before you begin {.tabset .tabset-fade .tabset-pills}

>Before the workshop you should have already downloaded [*GitHub Desktop*](https://desktop.github.com/) and created an [*GitHub*](https://github.com/) account. If you haven't already done so please do this **before** starting the tasks.

1. Open RStudio

2. Create a new R project for this workshop

3. Download the raw data for this workshop from wattle, and save it in a `/raw_data` file in the folder with your R project

4. Create a new ‘R Markdown’ file for the answers, give it a meaningful name, and save it in the folder with your R project

5. Delete the contents of the R markdown file, and replace it with the text below

6. Enter your U number and the date into the appropriate places in the top of your R markdown file

```
---
title: "[Name of this workshop] solutions"
author: "Your U Number"
date: "the date"
output:  
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: true
        toc_float: true
        highlight: pygments
---

# Task 1: 

# Task 2: 

# Task 3: 

# Task 4: 

```
<br>

#### Remember it's a lab book {.tabset .tabset-fade .tabset-pills}

Through this and every workshop, you should type as much descriptive text into your R Markdown document as someone (including but not limited to you) would need to understand what you're doing.

**Remember**: This document is your lab book! The information in it must be sufficient to reproduce your research, allow others to understand your way of thinking, and allow you to use it in your later work (e.g. when writing assessments, thesis, blogs or papers). Imagine that you are writing a manual and a report at the same time.

Please keep the level 1 headings we have given you above, so it's easy for us to come around and help you. Please use subheadings to divide up your work within each section, subheadings work like this:

```
## This is a level 2 subheading
### This is a level 3 subheading
```
<br>

# **Introduction**

Fundamental to project organisation is maintaining a consistent, uniform and simple set of working files that can be understood by your future self, collaborators and interested folks wanting to reproduce your analyses or re-use code or your data. We also want to keep a history of how such files change in case we find some new code we write results in older code no longer running. 

For complex projects with many collaborators maintaining tidiness and keeping track of changing files can be very challenging! While file sharing services like *Dropbox* and *OneDrive* are easy to use and can maintain a consistent file set on your computer(s) and/or among collaborators it has some disadvantages. It can’t tell you what has changed within files. It also can’t easily deal with conflicts between files when you and your collaborators are working at the same time on the same things. 

This often leads to a build up “too afraid to delete” files that no one, including yourself, knows what to do with, but you're too afraid to delete in case you ruin something. A little bit of organising can keep chaos from compounding and maintain reproducibility (for both you and others) through the development of data science projects.

Ideally, it would be nice to easily see the current (and hopefully best) set of files, but to be able to see who has changed them and what specific parts of the files have been modified. We also want a workflow that allows us to go back in time -- a history that has documented the evolution of files -- in case something has changed that breaks your code. Keeping a record of how files for a particular project has changed is also important to be able to quickly solve problems when (not if) they arise. 

### **Version Control to the Rescue!**

What is version control? Version control is a file storage, documentation and tracking system. In other words, it allows you to keep a documented history of files and file structures as a specific project develops. It's used extensively in data science for: 1) building code and programs; 2) analyzing biological data and 3) maintaining databases. 

>**Git** is a type of version control system that we'll be using in this workshop. It's extremely powerful and is probably the most widely used version control system, not only in biology, but a multitude of different fields (e.g. computer science, engineering etc). 

```{r, echo=FALSE}
embed_url("https://www.youtube.com/watch?v=xQujH0ElTUg") %>% use_align("center")
```

### **GitHub and GitHub Desktop**

Now that you know that *Git* is a type of version control system, and you know more about how version control works, what is *GitHub*? 

*Git* is great because it allows you to track files and file structure on your local computer. However, what if you want to collaborate? We need a 'cloud' interface that makes use of the power of *Git*, but allows everyone to have the same history of events. 

That's where *GitHub* comes into play. *GitHub* is essentially the cloud/online version of your *Git* project. It hosts your files, and the history of the project, in the cloud! *GitHub Desktop*, in contrast, is the application that you use to communicate changes on your local computer up to *GitHub*. Once changes to files in your repo are made on your local computer you then **add** these files to the staging area, **commit** a detailed message about what was changed and why, and then **push** these changes to *GitHub*. These changes are then available to be **pulled** down from the cloud by your collaborators. This allows a project to be in sync across many, many different local computers. 

```{r, fig.align='center', echo = FALSE, fig.cap="Git and GitHub make for a powerful collaborative version control system. GitHub Desktop is a useful application that is used to interface your local Git project with GitHub."}
magick::image_read("./videos/mgithub.jpg")
```


The **pull**--**add**--**commit**--**push** sequence is all your need to remember when working in a *GitHub* repository, particularly if you have collaborators. Before you start, **pull** down any new changes that might be in the cloud because your collaborator might have been pulling an 'all nighter' to get some data analysis done. You want to make sure that you have the most recent version of their files. Then, you go on to work normally and remember to **add** and **commit** files regularly as you get important work finished. You can push up all at once, or you may want to push changes a few times a day so that the project in the cloud has your most up-to-date files as well. 

>**Important**: Binary files like word and excel files are not good to track because Git will not be able to track their contents. Try to keep the files you track to simple tex-based files like `.txt` and `.csv`. 

#### Public versus Private Repositories

*GitHub* has two types of repositories. 'Private` repositories are those that are only accessed and viewed online by you and your collaborators on the project. 'Public' repositories are those that can be viewed by *anyone*, even if they are outside the project. 

By default, *GitHub* creates 'public' repos and you need to actually pay for 'private' repos. The reason is that *GitHub* is encouraging code sharing and transparency but also community coding by keeping the code open source. This is arguably one of the major successes of *GitHub*. Anyone can access the code, modify it, improve upon it or even update it with new features. This is incredibly valuable because it can result in code and data being more broadly accessible, usable and encourage new and exciting collaborations. 

Don't worry though. People can't just change your repo. They still need to be invited as collaborators or you need to approve of a `pull request` if these changes are to take effect in your repo. 

### **Advantages of Version Control (Git)** 
There are major advantages of using version control over just using *Dropbox* / *OneDrive*.

1. It keeps track of all files relevant to a specific project and/or that are necessary for its reproducibility by assigning them a unique identifier.

2. It documents the evolution of every file in a project, including data, manuscripts and code, by permitting the user to assign a commit tag describing how the file or set of files has changed.

3. It automatically indicates who, when and where a specific file was modified, identifying the in text locations that have changed.

4. It keeps a recorded history of all the changed files, allowing you to move back in time to look at older version of files or the state of the project at an earlier time.

5. It auto merges files that you and collaborators are working on at the same time and introduces conflict markers in the text of the file if the same lines have been edited. This forces you and others to deal with and correct conflicts immediately.

6. It makes running analyses and scripts on servers much easier.

7. It promotes easier collaboration between lab members and co-authors by keeping files in sync and annotated across diverse platforms.

8. It makes it much easier at the end of a project to quickly share and store data and code relevant for reproducing the project (analyses, paper etc) which is often a necessary pre-requisite for publishing.

9. It makes it easier to correct mistakes within a project by allowing you to track down the relevant colleague (or *vis versa*) when issues arise after updates or changes.

10. It forces you to think about the relevant parts of the project that are essential for it’s reproducibility and allows you to customize what is tracked.

# **Let's jump in!**

Now that you have some familiarity with *GitHub* and *Git*, lets try to get ourselves set up and orientated. In groups, find a partner and complete the following tasks. Think about how you would describe to others how to complete these tasks. Two of you from a single group will be chosen randomly to come up and teach the class how to complete each task.

# **Task 1**: Setting up a repository (repo)

Create a new repository (repo) on your [*GitHub*](https://github.com/) account. Keep the repo name: 1) informative; 2) short; 3) and avoid 'spaces' in your repo name. For point 3, you can us '_' or '.' between words. This makes it easier when you want to code path names to your project. 

# **Task 2**: Clone your repository

Great! You have a repo. Now what? It's currently up in the cloud. That's not as useful as having it on your computer. 

For your next task, clone down your repository to your computer. Cloning makes a 'copy' of your new repo -- creating a new folder on your computer. This folder is somewhat special though because it retains 'lines of communication' with the folder on *GitHub*. 

Use the [*GitHub Desktop*](https://desktop.github.com/) application you downloaded on your computer to clone your new repository. 

<br>

:::: {.greenbox data-latex=""}
::: {.center data-latex=""}
<font size="18">**Relax!**</font>
:::
**When you get to this point, put your green post-it on your laptop (green tick on zoom), and, relax. See if you can help others in your group. We'll have two reps from a random group present their solutions soon.**
::::

# **Task 3**: Adding files, commiting (staging) and pushing to GitHub

Now that you have your *GitHub* repo cloned (i.e., downloaded) onto your computer navigate to this folder on your computer. 

>**Hint**: `GitHub Desktop` has a handy feature on the repo's front page for you to click on and take you directly to the folder)

Create: 1) a new RStudio project file and 2) add a new Rmarkdown file in this directory. 

Once added go back to your *GitHub Desktop* application. What's changed? Commit a message about why you added these files and push this up to your *GitHub* repo.

# **Task 4**: Editing file, saving, committing and pushing to GitHub

Open up your new Rmarkdown file that you created and are now tracking in your repo. Edit this file by making your own summary notes (can just be bullet points on each step) on **Tasks 1 & 2** so you remember how to create a repo and clone this down on your computer. 

Save this file, commit and push it up to *GitHub*. 

:::: {.greenbox data-latex=""}
::: {.center data-latex=""}
<font size="18">**Relax!**</font>
:::
**When you get to this point, put your green post-it on your laptop (green tick on zoom), and, relax. See if you can help others in your group. We'll have two reps from a random group present their solutions soon.**
::::


# **Task 5**: Adding collaborators to your repo

You now have a repo that's ready to share with the world. You've made this repo public, so everyone can see it, which is very transparent! Others can always join in on the fun. Great work. 

Let's now invite a collaborator. Complete the following:

1) Team up with the person beside you. 
2) Ask them for their GitHub username and add them to your *GitHub* repo. This is done on your *GitHub* repo through the browser by managing collaborators in the `settings` tab. You should also give **them** your username so they can do the same for you. 
3) Once added to their repository, clone their repository onto your local computer. 
4) Open up their Rmarkdown file that describes the steps for creating and cloning a repo. Read these steps. Edit one of the steps that is unclear to you to try and make it more clear. 
5) Save the file. 
6) Commit and push these changes up.

# **Task 6**: Pulling down changes and creating an issue on *GitHub*

Your partner has made some changes. You won't see these changes yet because they are on *GitHub* in the cloud!

Complete the following steps:

1) Go to your *GitHub Desktop* application. Pull down changes to *your* repo that your partner sent to *GitHub*.
2) Review these changes. In the Rmarkdown header add the date these changes were finalised. Commit and push up these new changes.

Now that you have a collaborator it's a good idea to create a To Do list. One place to do this is though *GitHub* issues. Issues can be used to identify bugs in code or suggest enhancements / additions to code, to get collaborators to check data or even just create a to do list so everyone has tasks that they can keep track of. We'll use it to set up a set of tasks for the project. Assume the project will be for you and your collaborator to analyse a big biological dataset together. 

Complete the following steps:

1) Create a new 'To Do List' issue on your *GitHub* project.
2) Create a *To Do* 'tick' box that lists three different tasks that will be the next steps in setting up the project. These tasks could be anything. Adding data, updating the README or meta-data, running some exploratory analysis and adding plotting code. 
3) Use tags to identify who (you or your partner) is meant to do those tasks. 

:::: {.greenbox data-latex=""}
::: {.center data-latex=""}
<font size="18">**Relax!**</font>
:::
**When you get to this point, put your green post-it on your laptop (green tick on zoom), and, relax. See if you can help others in your group. We'll have two reps from a random group present their solutions soon.**
::::

# **Task 7**: Resolving conflicts

Conflicts are common. They occur when you and your collaborator are working on the same lines of the same file. If you are working on different parts of the same file Git will auto-merge these changes. It will do this seamlessly for you. However, it's more common than you think that you and your collaborator are editing the same place of a file. *Git* is powerful for resolving these conflicts (assuming you're working with .txt or .csv style files). 

Create a conflict yourself and learn how to recognize conflicts and resolve them when they happen. 

Complete the following:

1) You and your partner should open the same file at the same time. To keep it simple, open the `README` file. 
2) Each of you edit the same line in the file. For example, each of you can give the title of the project a different name.
3) Each of you save the file.
4) Tell your partner to commit and push up their changes to *GitHub*. You do the same with your partner's repo.
5) Each of you now pull down the change. What do you notice happens to your file when you pull down the changes? What does *GitHub Desktop* tell you?
6) Open the 'conflicted' README file.
7) What do you see in the file? Describe in your notebook what is different about the README file.
8) Resolve the conflict. Pick a change you like best, delete all other conflict markers and text.
9) Commit and push up the changes to *GitHub*

<br> 

:::: {.greenbox data-latex=""}
::: {.center data-latex=""}
<font size="18">**Relax!**</font>
:::
**When you get to this point, put your green post-it on your laptop (green tick on zoom), and, relax. See if you can help others in your group. We'll have two reps from a random group present their solutions soon.**
::::

<br>

:::: {.bluebox data-latex=""}
::: {.center data-latex=""}
<font size="18">**Stretch Tasks**</font>
:::
**The next couple of tasks are bonus. Try these out if we have time. These tasks will allow you to better familiarise yourself with *GitHub* and show you some handy features to make your repo more visible/shareable.**
::::

<br>

# **Task 8**: Branching and merging

If you're coding and want to do something experimental it's good practice to do this on a separate branch to make sure that your existing code doesn't break. This is very useful for analysis or building your own software. *GitHub* also makes use of certain branch names to create webpages that can be used to render and host your code and reports. *GitHub* generally designates this branch as `gh-pages`. 

Try creating a new branch in your repo. Call this branch `testing`. 


# **Task 9**: GitHub Webpage 

*GitHub* is not only a great place for hosting your project, but it's also useful for hosting webpages! This can be really useful for providing a simple way for others to 'see' your results. You simply just send them the link.
